---
title: "Querying FHIR R4 Patients via objectSetsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Querying FHIR R4 Patients via objectSetsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

FHIR (Fast Healthcare Interoperability Resources) R4 is the dominant standard
for healthcare data exchange. `ontologyConnectR` provides a native FHIR
connector that:

- Pushes `eq`, `contains`, `gt`, `gte`, `lt`, `lte`, and `in` filters directly
  to the FHIR search API — no client-side post-filtering needed.
- Follows FHIR Bundle pagination (`Bundle.link[relation=next]`) automatically.
- Evaluates FHIRPath expressions in `schema_map` to map nested fields to flat
  property columns.

## Creating a FHIR connector

```{r connector}
library(ontologyConnectR)

patient_conn <- conn_fhir(
  base_url      = "https://hapi.fhir.org/baseR4",
  resource_type = "Patient",
  schema_map    = list(
    patient_id   = "id",
    family_name  = "name.where(use='official').family",
    given_name   = "name.where(use='official').given.first()",
    birth_date   = "birthDate",
    gender       = "gender",
    deceased     = "deceased.ofType(boolean)"
  )
)
```

## Building the context

```{r context}
bundle <- list(
  object_types = list(
    Patient = list(id = "Patient", extensions = list())
  ),
  link_types = list()
)

ctx <- ontology_context(bundle, connectors = list(Patient = patient_conn))
```

## Querying with filter push-down

```{r query}
# Both filters are pushed to the FHIR server: ?birthdate=gt1990-01-01&gender=female
result <- object_set(ctx, "Patient") |>
  os_filter("birth_date", "gt", "1990-01-01") |>
  os_filter("gender", "eq", "female") |>
  os_limit(20L) |>
  os_collect()

head(result)
```

## Filter operator mapping

| objectSetsR op | FHIR search modifier | Example URL parameter         |
|----------------|----------------------|-------------------------------|
| `eq`           | exact match          | `?family=Smith`               |
| `contains`     | `:contains`          | `?family:contains=smi`        |
| `gt`           | `gt` prefix          | `?birthdate=gt1990-01-01`     |
| `gte`          | `ge` prefix          | `?birthdate=ge1990-01-01`     |
| `lt`           | `lt` prefix          | `?birthdate=lt2000-01-01`     |
| `lte`          | `le` prefix          | `?birthdate=le2000-01-01`     |
| `in`           | comma-separated      | `?status=active,inactive`     |

## FHIRPath schema maps

```{r fhirpath}
encounter_conn <- conn_fhir(
  base_url      = "https://hapi.fhir.org/baseR4",
  resource_type = "Encounter",
  schema_map    = list(
    encounter_id = "id",
    status       = "status",
    class_code   = "class.code",
    patient_ref  = "subject.reference",
    start_date   = "period.start",
    end_date     = "period.end"
  )
)
```

## Authentication

```{r auth}
secured_conn <- conn_fhir(
  base_url      = "https://fhir.nhs.uk/R4",
  resource_type = "Patient",
  schema_map    = list(patient_id = "id", family_name = "name[0].family"),
  auth          = conn_auth_bearer(env_var = "NHS_FHIR_TOKEN")
)
```

## Caching

Clinical reference data (e.g. code systems) rarely changes. Cache aggressively:

```{r cache}
# Cache for 1 hour
concept_conn <- conn_fhir(
  base_url      = "https://tx.fhir.org/r4",
  resource_type = "CodeSystem",
  schema_map    = list(id = "id", name = "name", url = "url"),
  cache_ttl     = 3600L
)
```
