---
title: "ICIJ Offshore Leaks: Graph Analysis with ontologyConnectR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ICIJ Offshore Leaks: Graph Analysis with ontologyConnectR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The ICIJ Offshore Leaks database exposes millions of records from the Panama
Papers, Pandora Papers, and other leaks. `ontologyConnectR` ships a bundled
sample and a `conn_icij()` connector that wraps it in the standard
OntologyConnector interface.

This vignette shows how to query entities, filter by jurisdiction, and load
`officer_of` edges to find the people behind shell companies.

## Loading the connectors

```{r setup-conn}
library(ontologyConnectR)

# Uses the bundled sample — no download needed
entity_conn      <- conn_icij("Entity")
officer_conn     <- conn_icij("Officer")
intermediary_conn <- conn_icij("Intermediary")

# Or point at your own ICIJ CSV directory:
# entity_conn <- conn_icij("Entity", csv_dir = "~/data/icij")
```

## Building the context

```{r context}
bundle <- list(
  object_types = list(
    Entity      = list(id = "Entity",      extensions = list()),
    Officer     = list(id = "Officer",     extensions = list()),
    Intermediary = list(id = "Intermediary", extensions = list())
  ),
  link_types = list()
)

ctx <- ontology_context(
  bundle     = bundle,
  connectors = list(
    Entity       = entity_conn,
    Officer      = officer_conn,
    Intermediary = intermediary_conn
  )
)
```

## Querying entities

```{r query-entities}
# All Panama Papers entities in the British Virgin Islands
bvi_panama <- object_set(ctx, "Entity") |>
  os_filter("source_id",    "eq", "Panama Papers") |>
  os_filter("jurisdiction", "eq", "BVI") |>
  os_collect()

nrow(bvi_panama)
head(bvi_panama[, c("name", "status", "incorporation_date")])
```

## Counting by filter

```{r count}
os_count(
  os_filter(object_set(ctx, "Entity"), "source_id", "eq", "Panama Papers")
)
```

## Inspecting the pending query

```{r show-query}
os <- object_set(ctx, "Entity") |>
  os_filter("jurisdiction", "eq", "PAN") |>
  os_limit(100L)

os_show_query(os)
# ConnectorObjectSet [Entity]
#   Connector: conn_icij
#   Pending filters (1):
#     jurisdiction eq PAN
#   Limit: 100
```

## Loading edges for graph analysis

The `icij_load_edges()` function loads the `edges.csv` file which records
relationships between entities, officers, and intermediaries:

```{r edges}
edges <- icij_load_edges()   # uses the bundled sample

# Find all officers connected to a specific entity
officer_of <- edges[edges$rel_type == "officer_of", ]
head(officer_of[, c("node_id_start", "node_id_end", "rel_type")])

# Join with officer data
officers <- conn_fetch(officer_conn)
linked   <- merge(
  officer_of,
  officers,
  by.x = "node_id_start",
  by.y = "node_id"
)
head(linked[, c("name", "node_id_end", "source_id")])
```

## ICIJ schema maps

Each entity type has a pre-built schema map. Key properties:

| Property ID              | Entity | Officer | Intermediary | Address |
|--------------------------|--------|---------|--------------|---------|
| `node_id`                | ✓      | ✓       | ✓            | ✓       |
| `name`                   | ✓      | ✓       | ✓            |         |
| `jurisdiction`           | ✓      |         |              |         |
| `status`                 | ✓      |         | ✓            |         |
| `source_id`              | ✓      | ✓       | ✓            | ✓       |
| `countries`              | ✓      | ✓       | ✓            | ✓       |
| `incorporation_date`     | ✓      |         |              |         |
| `address`                | ✓      |         | ✓            | ✓       |
