---
title: "Connecting a REST API as an Object Type"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connecting a REST API as an Object Type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Overview

`ontologyConnectR` allows you to back an ontology object type against a live
REST API without materialising data locally. Every `os_collect()` call
translates accumulated filters to HTTP query parameters and fetches results in
real time.

## Defining the connector

```{r connector}
library(ontologyConnectR)

patient_conn <- conn_rest(
  base_url    = "https://api.example-hospital.org/v2",
  list_path   = "/patients",
  get_path    = "/patients/{pk}",
  count_path  = "/patients/count",
  pk_property = "patient_id",
  schema_map  = list(
    patient_id  = "id",
    given_name  = "name[0].given[0]",
    family_name = "name[0].family",
    birth_date  = "birthDate",
    gender      = "gender"
  ),
  auth       = conn_auth_bearer(env_var = "HOSPITAL_API_TOKEN"),
  pagination = conn_pagination_link(next_path = "links.next"),
  cache_ttl  = 300L   # Cache results for 5 minutes
)
```

## Building the ontology context

```{r context}
bundle <- list(
  object_types = list(
    Patient = list(id = "Patient", extensions = list())
  ),
  link_types = list()
)

ctx <- ontology_context(
  bundle     = bundle,
  connectors = list(Patient = patient_conn)
)
```

## Querying

```{r query}
# Accumulate filters — no network call yet
os <- object_set(ctx, "Patient") |>
  os_filter("gender", "eq", "female") |>
  os_limit(50L)

# Show what will be fetched (human-readable, not SQL)
os_show_query(os)

# Execute — one HTTP call per page, pagination handled automatically
patients <- os_collect(os)
head(patients)
```

## Pagination

`conn_pagination_link()` follows RFC 5988 `Link: <url>; rel="next"` headers
automatically. For page-number APIs use `conn_pagination_offset()`:

```{r pagination}
paginated_conn <- conn_rest(
  base_url    = "https://api.example.com",
  list_path   = "/records",
  get_path    = "/records/{pk}",
  schema_map  = list(id = "id", name = "name"),
  pk_property = "id",
  pagination  = conn_pagination_offset(
    page_param = "page",
    size_param = "per_page",
    page_size  = 100L
  )
)
```

## Authentication

```{r auth}
# Bearer token (e.g. JWT)
auth_jwt <- conn_auth_bearer(env_var = "MY_JWT_TOKEN")

# HTTP Basic Auth
auth_basic <- conn_auth_basic(username = "alice",
                               password_env_var = "MY_PASSWORD")

# API Key in custom header
auth_key <- conn_auth_api_key(header = "X-API-Key",
                               key_env_var = "MY_API_KEY")

# OAuth2 client credentials (token refresh handled automatically)
auth_oauth <- conn_auth_oauth2(
  token_url             = "https://auth.example.com/oauth/token",
  client_id             = "my-client-id",
  client_secret_env_var = "MY_CLIENT_SECRET",
  scope                 = "read:patients"
)
```

## Caching

Set `cache_ttl` (seconds) on the connector to avoid redundant HTTP calls:

```{r caching}
conn <- conn_rest(
  base_url    = "https://api.example.com",
  list_path   = "/items",
  get_path    = "/items/{pk}",
  schema_map  = list(id = "id"),
  pk_property = "id",
  cache_ttl   = 300L   # 5-minute TTL
)

# Invalidate on demand (e.g. after a write operation)
conn_invalidate(conn)

# Change TTL without reconstructing the connector
conn <- conn_set_ttl(conn, 60L)
```
